Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification

### 2021-09-05 paper-revision/policar2020-tsne-embedding.tex
Add final paper revision

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{hyperref}

@@

\begin{document}
%
\title{Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}
%
\titlerunning{t-SNE Embedding and Batch Effects}
%
\author{Pavlin G. Poli\v{c}ar\inst{1} \and
Martin Stra\v{z}ar\inst{1} \and
Bla\v{z} Zupan\inst{1,2}}
%
\authorrunning{P. G. Poli\v{c}ar, M. Stra\v{z}ar, and B. Zupan}
%
\institute{University of Ljubljana, Faculty of Computer and Information Science, Večna pot 113, Ljubljana
\email{\{pavlin.policar,martin.strazar,blaz.zupan\}@fri.uni-lj.si}\\[2pt]
\and
Baylor College of Medicine, Houston, TX 77030, USA
}

@@

Due to their severity, numerous computational approaches have been proposed to
directly remove batch effects when performing joint analysis on two or more
data sets. Batch effect removal is typically performed as a preprocessing
step. Existing approaches involve either modifying the original data matrix
or finding a joint lower-dimensional space, where batch effects are removed.
Current methods broadly fall into two categories:
\begin{enumerate}
\item mutual nearest neighbor-based approaches aim to identify matching
 populations of cells across the data sets, using them to either find and
 correct the data sets~\cite{Haghverdi2018} or directly construct a
 batch-corrected k-nearest neighbor graph used in downstream analyses~\cite
 {Park2018}.
\item embedding multiple data sets into a joint lower-dimensional space, where
 batch effects are removed. Some of these approaches opt for linear
 dimensionality-reduction methods such as PCA~\cite{Korsunsky2019} or
 MultiCCA~\cite{Butler2018}, while others employ non-linear techniques from
 deep learning~\cite{Li2020,Lopez2018}. Still, other approaches use a
 combination of the two~\cite{Stuart2019,Hie2019}. Note that these approaches bear similarity with transfer learning~\cite{https://journalofbigdata.springeropen.com/articles/10.1186/s40537-016-0043-6}, which has also been used in domain adaptation~\cite{https://jieliu6.github.io/publication/conference-paper/jointly-embedding-multiple-single-cell/}.
\end{enumerate}

@@

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macosko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after prepossessing.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{Bard2005}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

@@

We use a number of additional, recently proposed modifications to enhance the
t-SNE visualization of the reference data set. Kobak and Linderman have shown
that the global consistency of embeddings produced by popular visualization
algorithms are largely dependent on their initialization~\cite
{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
more meaningful layouts of the resulting clusters (Fig.~\ref
{fig:multiscale}.b) as opposed to using randomly initialized embeddings
(Fig.~\ref{fig:multiscale}.a). Another important extension is the use of
multi-scale similarities, which, in addition to considering short range
interactions, also models wider point neighborhoods. Coupled with PCA-based
initialization, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref
{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic
neurons, corresponding to their developmental origin, as discussed in
Tasic \etal~\cite{Tasic2018}, while this division is less apparent when using
PCA-based initialization alone in Fig.~\ref{fig:multiscale}.b.

@@

Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based
positioning alone is insufficient. We construct a reference embedding using
only neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref
{fig:optimization}.a) and use that to position neuronal cells from the data
set from Campbell \etal~\cite{Campbell2017}. We utilize the weighted mean and
median positions to initialize point positions from the secondary data set,
as shown in Figs.~\ref{fig:optimization}.b and \ref
{fig:optimization}.c. After initialization, we optimize point positions using
the procedure described above for 500 iterations. The resulting
visualizations from both initializations are visually very similar,
indicating stable convergence. We show one of the resulting visualizations in
Fig.~\ref{fig:optimization}.b.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{optimization.pdf}
  \caption{Comparison of different initialization schemes for positioning new
   data points onto reference embeddings. {\bf (a)} We construct a reference
   embedding using only neuronal subtypes from Hrvatin \etal~\cite
   {Hrvatin2018}. {\bf (b)} We position neuronal cells from
   Campbell \etal~\cite{Campbell2017} using the median initialization scheme
   from Kobak and Berens~\cite{Kobak2019} and run optimization for 500
   iterations. Compare the optimized embedding with the initial median
   initialization {\bf (c)} or by using a simple weighted mean
   initialization {\bf (d)}.}
  \label{fig:optimization}
\end{figure}

@@

We illustrate this with Fig.~\ref{fig:incomplete_reference}, where we first fit
create a reference embedding containing only neuronal cells from Hrvatin
\etal~\cite{Hrvatin2018}. We then select only non-neuronal cells from Campbell
\etal~\cite{Campbell2017} and add them to the reference embedding in Fig.~\ref
{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal are
scattered somewhat arbitrarily around several clusters in the reference
embedding. Interestingly, the secondary data points form a ``ring'' around
one of the clusters, indicating that these data points are very different
from the cells in this cluster. Notice also that the points from the
secondary data set exhibit little to no clustering and the different cell
types seem to be mixed among each other. We hypothesize that this effect is
due primarily to the single-cell preprocessing pipeline and not the
limitations of our procedure itself, as the informative genes selected to
create the reference neuronal embedding likely do not differentiate
supportive glial cells from the secondary data set. This effect is similar
to procedures such as scMap-Cluster, a consensus $k$-nearest neighbor 
method, which heuristically determines a distance threshold to identify
unknown cell types~\cite{Kiselev2018}.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{incomplete_reference-brain.pdf}
  \caption{A reference embedding must contain all the cell types in the secondary embedding to produce reliable results. {\bf (a)} We construct a reference embedding containing only neuronal cells from Hrvatin \etal~\cite {Hrvatin2018}. {\bf (b)} We select only non-neuronal cells from Campbell \etal~\cite{Campbell2017} so that no overlap exists between the cell types between the data sets. Conceptually, t-SNE behaves similarly to a $k$-NN classifier and places the non-neuronal cells to their most similar points in the reference. In some instances, the non-neuronal cells are sufficiently different from the neuronal cells so that they are repelled from the reference clusters. Such behavior results in the ``ring'' seen on the right-hand side of the embedding.} \label{fig:incomplete_reference}
\end{figure}

@@

Single-cell data sets suffer from high levels of technical noise and low
capture efficiency, resulting in sparse and noisy expression matrices~\cite
{umi}. A common occurrence in these data sets is ``dropout'', where an
expressed gene is not measured, and its corresponding matrix entry is set to
zero. To address this problem, we use a specialized feature-selection method,
which exploits the mean-dropout relationship of expression counts as recently
proposed by Kobak and Berens~\cite{Kobak2019}. In this context, we will refer
to all genes with matrix entries set to zero as dropouts. Intuitively, if a
gene has high mean expression over all cells, but is detected in only a
handful of them (i.e. has high mean dropout rate), then this gene is likely
specific to a specific cell-type and will serve as a good feature for any
subsequent analysis where we wish to discriminate between cell types.

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{multirow}
\usepackage{booktabs}
\usepackage{hyperref}

@@

\begin{document}

%
\title{Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}
%
\titlerunning{t-SNE Embedding and Batch Effects}
%
\author{Pavlin G. Poli\v{c}ar\inst{1,}\thanks{Corresponding author} \and
Martin Stra\v{z}ar\inst{1} \and
Bla\v{z} Zupan\inst{1,2}}
%
\authorrunning{P. G. Poli\v{c}ar, M. Stra\v{z}ar, and B. Zupan}
%
\institute{University of Ljubljana, Faculty of Computer and Information Science, Večna pot 113, Ljubljana
\email{\{pavlin.policar,martin.strazar,blaz.zupan\}@fri.uni-lj.si}\\[2pt]
\and
Baylor College of Medicine, Houston, TX 77030, USA
}

@@

Due to their severity, numerous computational approaches have been proposed to
directly remove batch effects when performing joint analysis on two or more
data sets. Batch effect removal is typically performed as a preprocessing
step. Existing approaches involve either modifying the original data matrix
or finding a joint lower-dimensional space, where batch effects are removed.
Current methods broadly fall into two categories:
\begin{enumerate}
\item mutual nearest neighbor-based approaches aim to identify matching
 populations of cells across the data sets, using them to either find and correct the data sets~\cite{Haghverdi2018} or directly construct a
 batch-corrected k-nearest neighbor graph used in downstream analyses~\cite{Park2018}.
\item embedding multiple data sets into a joint lower-dimensional space, where
 batch effects are removed. Some of these approaches opt for linear
 dimensionality-reduction methods such as PCA~\cite{Korsunsky2019} or
 MultiCCA~\cite{Butler2018}, while others employ non-linear techniques from
 deep learning~\cite{Li2020,Lopez2018}. Still, other approaches use a
 combination of the two~\cite{Stuart2019,Hie2019}. Note that these approaches bear similarity with transfer learning~\cite{weiss2016survey}, which has also been used in domain adaptation~\cite{liu2019jointly}.
\end{enumerate}

@@

\vspace{-3mm}
\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. The first data set in each pair
(Hrvatin \etal, Baron \etal, and Macosko \etal) was used as a reference.  We
relied on the quality control and annotations from the original publication
and report the number of cell types after preprocessing.  The cell
annotations were made consistent to annotations from the Cell
Ontology~\cite{Bard2005}. Notice that different RNA sequencing protocols
were used to estimate gene expressions.}
\label{tab:data sets}
\end{table}

@@

We use a number of additional, recently proposed modifications to enhance the
t-SNE visualization of the reference data set. Kobak and Linderman have shown
that the global consistency of embeddings produced by popular visualization
algorithms are largely dependent on their initialization~\cite{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b) as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a). Another important extension is the use of
multi-scale similarities, which, in addition to considering short range
interactions, also models wider point neighborhoods. Coupled with PCA-based
initialization, this produces even more meaningful visualizations where
clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic
neurons, corresponding to their developmental origin, as discussed in
Tasic \etal~\cite{Tasic2018}, while this division is less apparent when using
PCA-based initialization alone in Fig.~\ref{fig:multiscale}.b.

@@

Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based
positioning alone is insufficient. We construct a reference embedding using
only neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref{fig:optimization}.a) and use that to position neuronal cells from the data
set from Campbell \etal~\cite{Campbell2017}. We utilize the weighted mean and
median positions to initialize point positions from the secondary data set,
as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c. After initialization, we optimize point positions using
the procedure described above for 500 iterations. The resulting
visualizations from both initializations are visually very similar,
indicating stable convergence. We show one of the resulting visualizations in
Fig.~\ref{fig:optimization}.b.

@@

\begin{figure}[htbp]
  \includegraphics[width=\textwidth]{optimization.pdf}
  \caption{Comparison of different initialization schemes for positioning new
   data points onto reference embeddings. {\bf (a)} We construct a reference
   embedding using only neuronal subtypes from Hrvatin \etal~\cite{Hrvatin2018}. {\bf (b)} We position neuronal cells from
   Campbell \etal~\cite{Campbell2017} using the median initialization scheme
   from Kobak and Berens~\cite{Kobak2019} and run optimization for 500
   iterations. Compare the optimized embedding with the initial median
   initialization {\bf (c)} or by using a simple weighted mean
   initialization {\bf (d)}.}
  \label{fig:optimization}
\end{figure}

@@

We illustrate this with Fig.~\ref{fig:incomplete_reference}, where we first fit
create a reference embedding containing only neuronal cells from Hrvatin
\etal~\cite{Hrvatin2018}. We then select only non-neuronal cells from Campbell
\etal~\cite{Campbell2017} and add them to the reference embedding in Fig.~\ref{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal are
scattered somewhat arbitrarily around several clusters in the reference
embedding. Interestingly, the secondary data points form a ``ring'' around
one of the clusters, indicating that these data points are very different
from the cells in this cluster. Notice also that the points from the
secondary data set exhibit little to no clustering and the different cell
types seem to be mixed among each other. We hypothesize that this effect is
due primarily to the single-cell preprocessing pipeline and not the
limitations of our procedure itself, as the informative genes selected to
create the reference neuronal embedding likely do not differentiate
supportive glial cells from the secondary data set. This effect is similar
to procedures such as scMap-Cluster, a consensus $k$-nearest neighbor 
method, which heuristically determines a distance threshold to identify
unknown cell types~\cite{Kiselev2018}.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{incomplete_reference-brain.pdf}
  \caption{A reference embedding must contain all the cell types in the secondary embedding to produce reliable results. {\bf (a)} We construct a reference embedding containing only neuronal cells from Hrvatin \etal~\cite{Hrvatin2018}. {\bf (b)} We select only non-neuronal cells from Campbell \etal~\cite{Campbell2017} so that no overlap exists between the cell types between the data sets. Conceptually, t-SNE behaves similarly to a $k$-NN classifier and places the non-neuronal cells to their most similar points in the reference. In some instances, the non-neuronal cells are sufficiently different from the neuronal cells so that they are repelled from the reference clusters. Such behavior results in the ``ring'' seen on the right-hand side of the embedding.} \label{fig:incomplete_reference}
\end{figure}

@@

Single-cell data sets suffer from high levels of technical noise and low
capture efficiency, resulting in sparse and noisy expression matrices~\cite{umi}. A common occurrence in these data sets is ``dropout'', where an
expressed gene is not measured, and its corresponding matrix entry is set to
zero. To address this problem, we use a specialized feature-selection method,
which exploits the mean-dropout relationship of expression counts as recently
proposed by Kobak and Berens~\cite{Kobak2019}. In this context, we will refer
to all genes with matrix entries set to zero as dropouts. Intuitively, if a
gene has high mean expression over all cells, but is detected in only a
handful of them (i.e. has high mean dropout rate), then this gene is likely
specific to a specific cell-type and will serve as a good feature for any
subsequent analysis where we wish to discriminate between cell types.
